Finally Tagless, Partially Evaluated

### 2008-09-12 tagless/tagless-final.tex
Rename x and y to e1 and e2 for consistency

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below is metacircular in that it
interprets each object\hyp language
operation as the corresponding metalanguage operation.
\ifshort
\begin{code3}
module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
  let int  (x:int)  = x         let bool (b:bool) = b
  let lam  f        = f         let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee () end
\end{code3}
\else
\begin{code}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)

  let int  (x:int)  = x
  let bool (b:bool) = b
  let lam  f        = f
  let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee ()
end
\end{code}
\fi
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding}
with~|R|\ifshort: \texttt{module EXR = EX(R)}\fi.
\ifshort\else
\begin{code}
module EXR = EX(R)
\end{code}
\fi
Thus, |EXR.test1 ()| evaluates to the untagged boolean value |true|.
\begin{comment}
%% commenting out this Haskell in JFP version as well
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
\end{comment}
It is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
The soundness of the object language's type system with respect to
the dynamic semantics specified by a definitional interpreter
follows from the soundness of the metalanguage's type system.
\ifshort
(The source code shows a total interpreter |L| that measures
the size of each object term.)
\fi
\ifshort
We can also generalize from~|R| to all interpreters;
these propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\begin{proposition}
If a closed object term~$e$ has type~$t$,
and the OCaml module |I| implements the signature |Symantics|,
then under the OCaml module definition
\[
    \texttt{\begin{tabular}{@{}l@{}}
    module RESULT = \\
    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
    \quad (I)
    \end{tabular}}
\]
evaluating the expression \texttt{RESULT.term ()} never gets stuck: it either
does not terminate or evaluates to a value of type
\texttt{('c,\:$M(t)$)\;I.repr} (polymorphic over~|'c|).
\end{proposition}
\begin{proof}
By Corollary~\ref{cor:typing} and the type soundness of (this fragment of) OCaml.
\end{proof}
\begin{corollary}
If a closed object term~$e$ has type~$t$,
then under the OCaml module definition
\[
    \texttt{\begin{tabular}{@{}l@{}}
    module RESULT = \\
    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
    \quad (R)
    \end{tabular}}
\]
evaluating the expression \texttt{RESULT.term ()} never gets stuck: it either
does not terminate or evaluates to a value of type~$M(t)$.
\end{corollary}
\ifshort\else
For variety, we show another interpreter, which measures the \emph{size}
(or \emph{length}, thus the mnemonic |L|)
of each object term, defined as the number of term
constructors.
\begin{code}
module L = struct
  type ('c,'dv) repr = int

  let int  (x:int)  = 1
  let bool (b:bool) = 1
  let lam  f        = f 0 + 1
  let app  e1 e2    = e1 + e2 + 1
  let fix  f        = f 0 + 1
  let add  e1 e2    = e1 + e2 + 1
  let mul  e1 e2    = e1 + e2 + 1
  let leq  x y      = x + y + 1
  let if_  eb et ee = eb + et () + ee () + 1
end
\end{code}
Now the OCaml expression
|let module E = EX(L) in E.test1 ()|
evaluates to |3|. This interpreter is not only tagless but also
total. It ``evaluates'' even seemingly divergent terms; for instance,
|app (fix (fun self -> self)) (int 1)| evaluates to $3$.
\fi

@@

% this code does not have the 'sv parameter. It shows up later.
\ifshort
\begin{SaveVerbatim}{5a}
module C = struct type ('c,'dv) repr = ('c,'dv) code
  let int (x:int)  = .<x>.            let bool (b:bool) = .<b>.
  let lam f        = .<fun x -> .~(f .<x>.)>.
  let app e1 e2    = .<.~e1 .~e2>.
  let fix f =  .<let rec self n = .~(f .<self>.) n in self>.
  let add e1 e2    = .<.~e1 + .~e2>.  let mul e1 e2 = .<.~e1 * .~e2>.
  let leq x y      = .<.~x <= .~y>.
  let if_ eb et ee = .<if .~eb then .~(et ()) else .~(ee ())>. end
\end{SaveVerbatim}
\else
\begin{SaveVerbatim}{5a}
module C = struct
  type ('c,'dv) repr = ('c,'dv) code

  let int (x:int)   = .<x>.
  let bool (b:bool) = .<b>.
  let lam f         = .<fun x -> .~(f .<x>.)>.
  let app e1 e2     = .<.~e1 .~e2>.
  let fix f         = .<let rec self n = .~(f .<self>.) n in self>.
  let add e1 e2     = .<.~e1 + .~e2>.
  let mul e1 e2     = .<.~e1 * .~e2>.
  let leq x y       = .<.~x <= .~y>.
  let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>.
end
\end{SaveVerbatim}
\fi
\begin{SaveVerbatim}{5b}
let module E = EX(C) in E.test1 ()
\end{SaveVerbatim}
\begin{SaveVerbatim}{5c}
let module E = EX(C) in E.testpowfix7
\end{SaveVerbatim}
\begin{SaveVerbatim}{5d}
.<fun x_1 -> (fun x_2 -> let rec self_3 = fun n_4 ->
   (fun x_5 -> if x_5 <= 0 then 1 else x_2 * self_3 (x_5 + (-1)))
   n_4 in self_3) x_1 7>.
\end{SaveVerbatim}

>>>>

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below is metacircular in that it
interprets each object\hyp language
operation as the corresponding metalanguage operation.
\ifshort
\begin{code3}
module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
  let int  (x:int)  = x         let bool (b:bool) = b
  let lam  f        = f         let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
  let leq  e1 e2    = e1 <= e2
  let if_  eb et ee = if eb then et () else ee () end
\end{code3}
\else
\begin{code}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)

  let int  (x:int)  = x
  let bool (b:bool) = b
  let lam  f        = f
  let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  e1 e2    = e1 <= e2
  let if_  eb et ee = if eb then et () else ee ()
end
\end{code}
\fi
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding}
with~|R|\ifshort: \texttt{module EXR = EX(R)}\fi.
\ifshort\else
\begin{code}
module EXR = EX(R)
\end{code}
\fi
Thus, |EXR.test1 ()| evaluates to the untagged boolean value |true|.
\begin{comment}
%% commenting out this Haskell in JFP version as well
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
\end{comment}
It is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
The soundness of the object language's type system with respect to
the dynamic semantics specified by a definitional interpreter
follows from the soundness of the metalanguage's type system.
\ifshort
(The source code shows a total interpreter |L| that measures
the size of each object term.)
\fi
\ifshort
We can also generalize from~|R| to all interpreters;
these propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\begin{proposition}
If a closed object term~$e$ has type~$t$,
and the OCaml module |I| implements the signature |Symantics|,
then under the OCaml module definition
\[
    \texttt{\begin{tabular}{@{}l@{}}
    module RESULT = \\
    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
    \quad (I)
    \end{tabular}}
\]
evaluating the expression \texttt{RESULT.term ()} never gets stuck: it either
does not terminate or evaluates to a value of type
\texttt{('c,\:$M(t)$)\;I.repr} (polymorphic over~|'c|).
\end{proposition}
\begin{proof}
By Corollary~\ref{cor:typing} and the type soundness of (this fragment of) OCaml.
\end{proof}
\begin{corollary}
If a closed object term~$e$ has type~$t$,
then under the OCaml module definition
\[
    \texttt{\begin{tabular}{@{}l@{}}
    module RESULT = \\
    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
    \quad (R)
    \end{tabular}}
\]
evaluating the expression \texttt{RESULT.term ()} never gets stuck: it either
does not terminate or evaluates to a value of type~$M(t)$.
\end{corollary}
\ifshort\else
For variety, we show another interpreter, which measures the \emph{size}
(or \emph{length}, thus the mnemonic |L|)
of each object term, defined as the number of term
constructors.
\begin{code}
module L = struct
  type ('c,'dv) repr = int

  let int  (x:int)  = 1
  let bool (b:bool) = 1
  let lam  f        = f 0 + 1
  let app  e1 e2    = e1 + e2 + 1
  let fix  f        = f 0 + 1
  let add  e1 e2    = e1 + e2 + 1
  let mul  e1 e2    = e1 + e2 + 1
  let leq  e1 e2    = e1 + e2 + 1
  let if_  eb et ee = eb + et () + ee () + 1
end
\end{code}
Now the OCaml expression
|let module E = EX(L) in E.test1 ()|
evaluates to |3|. This interpreter is not only tagless but also
total. It ``evaluates'' even seemingly divergent terms; for instance,
|app (fix (fun self -> self)) (int 1)| evaluates to $3$.
\fi

@@

% this code does not have the 'sv parameter. It shows up later.
\ifshort
\begin{SaveVerbatim}{5a}
module C = struct type ('c,'dv) repr = ('c,'dv) code
  let int (x:int)  = .<x>.            let bool (b:bool) = .<b>.
  let lam f        = .<fun x -> .~(f .<x>.)>.
  let app e1 e2    = .<.~e1 .~e2>.
  let fix f =  .<let rec self n = .~(f .<self>.) n in self>.
  let add e1 e2    = .<.~e1 + .~e2>.  let mul e1 e2 = .<.~e1 * .~e2>.
  let leq e1 e2    = .<.~e1 <= .~e2>.
  let if_ eb et ee = .<if .~eb then .~(et ()) else .~(ee ())>. end
\end{SaveVerbatim}
\else
\begin{SaveVerbatim}{5a}
module C = struct
  type ('c,'dv) repr = ('c,'dv) code

  let int (x:int)   = .<x>.
  let bool (b:bool) = .<b>.
  let lam f         = .<fun x -> .~(f .<x>.)>.
  let app e1 e2     = .<.~e1 .~e2>.
  let fix f         = .<let rec self n = .~(f .<self>.) n in self>.
  let add e1 e2     = .<.~e1 + .~e2>.
  let mul e1 e2     = .<.~e1 * .~e2>.
  let leq e1 e2     = .<.~e1 <= .~e2>.
  let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>.
end
\end{SaveVerbatim}
\fi
\begin{SaveVerbatim}{5b}
let module E = EX(C) in E.test1 ()
\end{SaveVerbatim}
\begin{SaveVerbatim}{5c}
let module E = EX(C) in E.testpowfix7
\end{SaveVerbatim}
\begin{SaveVerbatim}{5d}
.<fun x_1 -> (fun x_2 -> let rec self_3 = fun n_4 ->
   (fun x_5 -> if x_5 <= 0 then 1 else x_2 * self_3 (x_5 + (-1)))
   n_4 in self_3) x_1 7>.
\end{SaveVerbatim}
